### 2014-10-25 thesis/litreview/litreview.tex
Updated title of literature review
Update case of literature review headings

\chapter{Literature Review} \label{chap:litreview}

@@

\section{Statistical methods}
\subsection{Regression}
As we stated earlier, logistic regression is a commonly-used modelling
technique for prediction in the medical domain \citep{Tu1996}, and especially
in the trauma data analysis \citep{Gabbe2005}. A series of
predictor (independent) variables (or features) determine the probability of a
particular outcome by an equation of the form
\begin{equation}
\mathrm{log}\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1X_1 + \beta_2X_2 + \dots + \beta_nX_n
\end{equation}
where $p$ is the probability of the desired outcome, $\beta_i$ are coefficients
associated with each variable, and $X_i$ are the values of the independent
variables \citep{Hosmer1989}. A stepwise selection algorithm is often used to
choose which variables to include in the model out of all available variables:
different algorithms produce different models which are then compared against
one another to find the best model. Usually, a standard maximum-likelihood
procedure is followed in order to obtain values for the $\beta_i$'s, and models
can be tested using measures such as the area under the receiver operating
characteristic (ROC) curve (AUC) and the Hosmer-Lemeshow statistic.

@@

\subsection{Markov models}
There have been a number of studies attempting to model the length of stay
using discrete-time Markov processes. Kapadia et al. constructed a transition
probability matrix using 4 states for a pediatric intensive care unit (ICU)
application \citep{Kapadia2000}:
\begin{center}
\begin{tabular}{l|llll}
& Low & Medium & High & Exit \\
\hline
Low & 0.309 & 0.223 & 0.024 & 0.444 \\
Medium & 0.137 & 0.513 & 0.096 & 0.254 \\
High & 0.086 & 0.364 & 0.379 & 0.171 \\
Exit & 0 & 0 & 0 & 1 \\
\end{tabular}
\end{center}
From the matrix, we can say that the probability of transitioning from state
`Medium' to state `Low' is 0.137. Note that each row sums to 1, and that `Exit'
is an absorbing state: once entered, one can only re-enter the `Exit' state
with 100\% certainty. This represents patients leaving the ICU. How did they
obtain these probabilities? From their data set, they classified each patient
with an appropriate state (Low, Medium, High) corresponding to the severity of
their illness based on a particular mortality score (PRISM), and did this daily
over a few months; this gave them patient records that were simply strings of
L, M and H. To calculate the probability of transitioning from `Medium' to
`Low', they counted all occurrences of `Medium' to `Low' transitions and
divided this by the number of all transitions out of `Medium', and repeated
this for every transition. Using the theory of Markov processes, they are able
to calculate the average time taken before a person enters the `Exit' state,
which is precisely the length of stay. They found that the Markov model fitted
the data well and produced predictions that were very close to the observed
results in the test set \citep{Kapadia2000}, indicating that using such models
justifies further investigation.

@@

\subsection{Other statistical techniques}
A number of other techniques have been used in length of stay prediction, which
we will describe briefly.
Yau et al. \citep{Yau2003} proposed a \textit{mixture regression model} in
modelling
length of stay to overcome the problem of correlated observations in hospital
data, which allowed them to predict neonatal length of stay across heterogenous
sub-groups (short-stay and long-stay). The idea is that these two sub-groups
have different length of stay patterns, and the length of stay of any patient
is just the probability of them being in one of these two groups (as well as
taking into account the underlying distribution of the group itself). Applying
this method to a set of neonatal length of stay data, they were able to
identify which input factors contributed to which component (short or long
stay) and derive useful insight for hospitals.

@@

\section{Data mining methods}
Data mining is the analysis of (often large) observational data sets to find
unsuspected relationships and to summarise the data in novel ways that are both
understandable and useful to the data owner \citep{Hand2001}. Compared to
traditional statistical, it is a relatively recent field that emerged in the
1990s, and has
given researchers a variety of new tools to analyse and understand biomedical
datasets \citep{Bellazzi2008,Yoo2012}. Bellazzi \& Zupan \citep{Bellazzi2008}
have written a
comprehensive review paper of data mining techniques in clinical prediction
settings, and we refer the reader to their article for a more detailed
treatment of the techniques and their applications. Here we will describe the
major methods in the context of their application to predicting hospital length
of stay.

\subsection{Artificial neural networks}
Artificial neural networks have been widely used in clinical prediction
following their use in diagnosing acute myocardial infarction and various
applications in radiology \citep{Baxt1995}.
One of the earliest applications of neural networks to
length of stay prediction was by Tu \& Guerriere, who used a neural network
with the following structure \citep{Tu1993} to predict length of stay in
intensive care following cardiac surgery:
\begin{figure}[h]
\caption{}
\centering
\includegraphics[scale=0.95]{images/a2-nn}
\end{figure}

@@

\subsection{Extensions to the artificial neural network model}
Buchman et al. published a comparison of the performance of logistic regression
and neural networks on predicting whether or not patients would stay greater
than 7 days in a surgical intensive care unit \citep{Buchman1994}. Instead of
using one type of training algorithm as Tu \& Guerriere did, they obtained
results from three different neural network algorithms: backpropagation (the
standard against which other training methods are compared), probabilistic, and
generalised regression. Probabilistic neural networks are trained by a single
pass over the training data and use Bayes strategies and Parzen windows to
estimate a probability density function for prediction. Generalised regression
neural networks allow the definition of arbitrarily complex regression surfaces
which allows for greater modelling power and is able to produce continuous
values instead of simply a binary classification \citep{Sprecht1991}. They found
that all three neural networks performed noticeably better than the logistic
regression model, but that they all achieved fairly similar results (by the ROC
test). However, the AUC for the four models were not given.

@@

\subsection{Other data mining techniques}
As discussed, artificial neural networks have been used extensively in
predicting patient length of stay. However, a few other data mining techniques
have also been investigated, namely tree-based methods and support vector
machines.

>>>>

\chapter{Predicting the LOS: A Review} \label{chap:litreview}

@@

\section{Statistical Methods}
\subsection{Regression}
As we stated earlier, logistic regression is a commonly-used modelling
technique for prediction in the medical domain \citep{Tu1996}, and especially
in the trauma data analysis \citep{Gabbe2005}. A series of
predictor (independent) variables (or features) determine the probability of a
particular outcome by an equation of the form
\begin{equation}
\mathrm{log}\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1X_1 + \beta_2X_2 + \dots + \beta_nX_n
\end{equation}
where $p$ is the probability of the desired outcome, $\beta_i$ are coefficients
associated with each variable, and $X_i$ are the values of the independent
variables \citep{Hosmer1989}. A stepwise selection algorithm is often used to
choose which variables to include in the model out of all available variables:
different algorithms produce different models which are then compared against
one another to find the best model. Usually, a standard maximum-likelihood
procedure is followed in order to obtain values for the $\beta_i$'s, and models
can be tested using measures such as the area under the receiver operating
characteristic (ROC) curve (AUC) and the Hosmer-Lemeshow statistic.

@@

\subsection{Markov Models}
There have been a number of studies attempting to model the length of stay
using discrete-time Markov processes. Kapadia et al. constructed a transition
probability matrix using 4 states for a pediatric intensive care unit (ICU)
application \citep{Kapadia2000}:
\begin{center}
\begin{tabular}{l|llll}
& Low & Medium & High & Exit \\
\hline
Low & 0.309 & 0.223 & 0.024 & 0.444 \\
Medium & 0.137 & 0.513 & 0.096 & 0.254 \\
High & 0.086 & 0.364 & 0.379 & 0.171 \\
Exit & 0 & 0 & 0 & 1 \\
\end{tabular}
\end{center}
From the matrix, we can say that the probability of transitioning from state
`Medium' to state `Low' is 0.137. Note that each row sums to 1, and that `Exit'
is an absorbing state: once entered, one can only re-enter the `Exit' state
with 100\% certainty. This represents patients leaving the ICU. How did they
obtain these probabilities? From their data set, they classified each patient
with an appropriate state (Low, Medium, High) corresponding to the severity of
their illness based on a particular mortality score (PRISM), and did this daily
over a few months; this gave them patient records that were simply strings of
L, M and H. To calculate the probability of transitioning from `Medium' to
`Low', they counted all occurrences of `Medium' to `Low' transitions and
divided this by the number of all transitions out of `Medium', and repeated
this for every transition. Using the theory of Markov processes, they are able
to calculate the average time taken before a person enters the `Exit' state,
which is precisely the length of stay. They found that the Markov model fitted
the data well and produced predictions that were very close to the observed
results in the test set \citep{Kapadia2000}, indicating that using such models
justifies further investigation.

@@

\subsection{Other Statistical Techniques}
A number of other techniques have been used in length of stay prediction, which
we will describe briefly.
Yau et al. \citep{Yau2003} proposed a \textit{mixture regression model} in
modelling
length of stay to overcome the problem of correlated observations in hospital
data, which allowed them to predict neonatal length of stay across heterogenous
sub-groups (short-stay and long-stay). The idea is that these two sub-groups
have different length of stay patterns, and the length of stay of any patient
is just the probability of them being in one of these two groups (as well as
taking into account the underlying distribution of the group itself). Applying
this method to a set of neonatal length of stay data, they were able to
identify which input factors contributed to which component (short or long
stay) and derive useful insight for hospitals.

@@

\section{Data Mining Methods}
Data mining is the analysis of (often large) observational data sets to find
unsuspected relationships and to summarise the data in novel ways that are both
understandable and useful to the data owner \citep{Hand2001}. Compared to
traditional statistical, it is a relatively recent field that emerged in the
1990s, and has
given researchers a variety of new tools to analyse and understand biomedical
datasets \citep{Bellazzi2008,Yoo2012}. Bellazzi \& Zupan \citep{Bellazzi2008}
have written a
comprehensive review paper of data mining techniques in clinical prediction
settings, and we refer the reader to their article for a more detailed
treatment of the techniques and their applications. Here we will describe the
major methods in the context of their application to predicting hospital length
of stay.

\subsection{Artificial Neural Networks}
Artificial neural networks have been widely used in clinical prediction
following their use in diagnosing acute myocardial infarction and various
applications in radiology \citep{Baxt1995}.
One of the earliest applications of neural networks to
length of stay prediction was by Tu \& Guerriere, who used a neural network
with the following structure \citep{Tu1993} to predict length of stay in
intensive care following cardiac surgery:
\begin{figure}[h]
\caption{}
\centering
\includegraphics[scale=0.95]{images/a2-nn}
\end{figure}

@@

\subsection{Extensions to the Artificial Neural Network Model}
Buchman et al. published a comparison of the performance of logistic regression
and neural networks on predicting whether or not patients would stay greater
than 7 days in a surgical intensive care unit \citep{Buchman1994}. Instead of
using one type of training algorithm as Tu \& Guerriere did, they obtained
results from three different neural network algorithms: backpropagation (the
standard against which other training methods are compared), probabilistic, and
generalised regression. Probabilistic neural networks are trained by a single
pass over the training data and use Bayes strategies and Parzen windows to
estimate a probability density function for prediction. Generalised regression
neural networks allow the definition of arbitrarily complex regression surfaces
which allows for greater modelling power and is able to produce continuous
values instead of simply a binary classification \citep{Sprecht1991}. They found
that all three neural networks performed noticeably better than the logistic
regression model, but that they all achieved fairly similar results (by the ROC
test). However, the AUC for the four models were not given.

@@

\subsection{Other Data Mining Techniques}
As discussed, artificial neural networks have been used extensively in
predicting patient length of stay. However, a few other data mining techniques
have also been investigated, namely tree-based methods and support vector
machines.
